Interpreting Random Forest Predictions for Firearm Identification Using LIME

### 2018-11-19 paper/Interpreting Random Forest Predictions for Firearm Identification Using LIME.Rmd
fixed code so it now runs with the dataset with multiple implementations of lime

Suppose that a complex model has been fit on a training dataset with $k$ features and a response variable that may either be numeric or categorical. This trained model will be applied to a testing dataset containing the $k$ features to obtain predictions. It is of interest to understand which variables played an important role in the predictions. LIME will be applied to one of the cases in the testing dataset to obtain an explanation. To create the explanation, LIME starts by approximating the distribution of each of the $k$ features in the training dataset. It then simulates a large dataframe from these distributions. The simulated data are used to perform feature selection to determine the $m$ most important features. Then a weighted ridge regression model is fit to standardized versions of the selected $m$ features from the simulated data. The model assigns higher weights to observations closer to the observed case of interest. A linear model is used since it has a well known form and coefficients that can be interpreted. The coefficients from the model are extracted and used to "explain" the importance of the variables. Features with a larger absolute value of their coefficient are implied to play a larger role in the creation of the prediction.

@@

  # Specify the number of bins to use
  nbins = 2:10

  # Run the lime functions for the specified number of bins 
  hamby224_all_results <- map(.x = nbins, 
             .f = run_lime, # run_lime is one of my helper functions
             train = hamby173and252_train %>% select(rf_features),
             test = hamby224_test %>% arrange(case) %>% select(rf_features) %>% na.omit(),
             rfmodel = as_classifier(rtrees),
             label = "TRUE",
             nfeatures = 3)

@@

  # Save the lime and explain results objects
  saveRDS(hamby224_lime, "../data/hamby224_lime.rds")
  saveRDS(hamby224_explain, "../data/hamby224_explain.rds")

@@

  # Load in the lime results object
  hamby224_lime <- readRDS("../data/hamby224_lime.rds")
  hamby224_explain <- readRDS("../data/hamby224_explain.rds")

@@

  # Join the data and the explanations and edit and add additional variables
  # Create the feature bin labels using my function "bin_labeller"
  hamby224_test_explain <- hamby224_test %>%
    mutate(case = as.character(case)) %>%
    full_join(hamby224_explain_results, by = "case") %>%
    mutate(case = factor(case),
           feature = factor(feature),
           feature_desc = factor(feature_desc)) %>%
    mutate(feature_number = readr::parse_number(feature_desc),
           strictly_less = FALSE) %>%
    arrange(nbins)

  # mutate(feature_bin = mapply(bin_labeller, feature = feature, 
  #                               feature_value = feature_value))

@@

  # Reorder the variables of feature_desc and feature_bin for plotting purposes
  hamby224_test_explain <- hamby224_test_explain %>%
    mutate(feature_desc = reorder(feature_desc, strictly_less),
           feature_desc = reorder(feature_desc, feature_number),
           feature_desc = reorder(feature_desc, as.numeric(feature))) %>%
    select(case:feature_desc, feature_bin:strictly_less, data, prediction)

@@

```{r}
check <- hamby224_test_explain %>%
  select(-data, -prediction) %>%
  group_by(case) %>%
  slice(1) %>%
  ungroup() %>%
  select(case, set, rf_features, rfscore, model_prediction, model_r2) %>%
  mutate(diff = rfscore - model_prediction,
         mean = (rfscore + model_prediction) / 2)

sum((subset(check$diff, check$set == "Set 1"))^2, na.rm = TRUE)
sum((subset(check$diff, check$set == "Set 11"))^2, na.rm = TRUE)

check %>%
  filter(set == "Set 1") %>%
  select(rf_features, diff, -case) %>%
  gather(key = feature, value = feature_value, rf_features) %>%
  select(feature, feature_value, diff) %>%
  ggplot(aes(x = feature_value, y = diff)) + 
  geom_point() + 
  facet_wrap( ~ feature, scales = "free") + 
  geom_hline(yintercept = 0, color = "blue")
```

@@

```{r fig.width = 8, fig.height = 4}
ggplot(check, aes(x = rfscore, y = model_prediction)) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid( ~ set) + 
  theme_bw() + 
  stat_smooth(method = "lm")
```

@@

```{r fig.width = 8, fig.height = 4}
ggplot(check, aes(x = rfscore, y = abs(diff))) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid( ~ set) +
  theme_bw()
```

@@

```{r fig.width = 8, fig.height = 4}
ggplot(check, aes(x = rfscore, y = diff)) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid( ~ set) +
  geom_hline(yintercept = 0, color = "blue") +
  theme_bw()
```

@@

```{r}
hamby224_test_explain %>%
  ggplot(aes(x = feature_desc)) + 
  geom_bar() +
  coord_flip()
```

```{r}
hamby224_test_explain %>%
  filter(set == "Set 1") %>%
  mutate(rfscore = round(rfscore, 3)) %>%
  select(case, bullet1, bullet2, land1, land2, rfscore) %>%
  distinct() %>%
  ggplot(aes(x = land1, y = land2, label = bullet1, label2 = bullet2,
             text = paste('Bullets Compared: ', bullet1, "-", land1, 
                          "vs", bullet2, "-", land2,
                          '\nRandom Forest Score: ', 
                          ifelse(is.na(rfscore), "Missing due to tank rash", rfscore)))) +
  geom_tile(aes(fill = rfscore)) +
  facet_grid(bullet2 ~ bullet1, scales = "free") +
  theme_minimal() +
  scale_fill_gradient2(low = "darkgrey", high = "darkorange", midpoint = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "RF Score")
```

>>>>

Suppose that a complex model has been fit on a training dataset with $k$ features and a response variable that may either be numeric or categorical. This trained model will be applied to a testing dataset containing the $k$ features to obtain predictions. It is of interest to understand which variables played an important role in the predictions. LIME will be applied to one of the cases in the testing dataset to obtain an explanation. To create the explanation, LIME starts by approximating the distribution of each of the $k$ features in the training dataset. It then simulates a large data frame from these distributions. The simulated data are used to perform feature selection to determine the $m$ most important features. Then a weighted ridge regression model is fit to standardized versions of the selected $m$ features from the simulated data. The model assigns higher weights to observations closer to the observed case of interest. A linear model is used since it has a well known form and coefficients that can be interpreted. The coefficients from the model are extracted and used to "explain" the importance of the variables. Features with a larger absolute value of their coefficient are implied to play a larger role in the creation of the prediction.

@@

  # Specify the options to use with lime 
  hamby224_lime_cases <- data.frame(bin_continuous = rep(TRUE, 18),
                                    quantile_bins = c(rep(TRUE, 9), rep(FALSE, 9)),
                                    nbins = rep(2:10, 2),
                                    use_density = rep(NA, 18)) %>%
  mutate(case = 1:length(nbins)) %>%
  select(case, bin_continuous, nbins, quantile_bins, use_density)

  # Run the lime functions for the specified number of bins
  hamby224_all_results <- pmap(.l = as.list(hamby224_lime_cases[1,-1]),
             .f = run_lime, # run_lime is one of my helper functions
             train = hamby173and252_train %>% select(rf_features),
             test = hamby224_test %>% arrange(case) %>% select(rf_features) %>% na.omit(),
             rfmodel = as_classifier(rtrees),
             label = "TRUE",
             nfeatures = 3)

@@

  # Name the items in the lime list
  names(hamby224_lime) <- map_chr(1:dim(hamby224_lime_cases[1,])[1], function(case) 
      sprintf("case: bin_continuous = %s, quantile_bins = %s, nbins = %0.f, use_density = %s",
              hamby224_lime_cases$bin_continuous[case],
              hamby224_lime_cases$quantile_bins[case],
              hamby224_lime_cases$nbins[case],
              hamby224_lime_cases$use_density[case]))

  # Save the lime and explain results objects and the case info
  saveRDS(hamby224_lime, "../data/hamby224_lime.rds")
  saveRDS(hamby224_explain, "../data/hamby224_explain.rds")
  saveRDS(hamby224_lime_cases, "../data/hamby224_lime_cases.rds")

@@

  # Load in the lime results object
  hamby224_lime <- readRDS("../data/hamby224_lime.rds")
  hamby224_explain <- readRDS("../data/hamby224_explain.rds")
  hamby224_lime_cases <- readRDS("../data/hamby224_lime_cases.rds")

@@

  # Join the data and the explanations and edit and add additional variables
  # Create the feature bin labels using my function "bin_labeller"
  hamby224_test_explain <- hamby224_test %>%
    mutate(case = as.character(case)) %>%
    full_join(hamby224_explain, by = "case") %>%
    mutate(case = factor(case),
           feature_desc = factor(feature_desc),
           feature_bin = pmap_chr(list(feature = feature, 
                                 feature_value = feature_value,
                                 b_c = bin_continuous,
                                 q_b = quantile_bins,
                                 n_b = nbins,
                                 u_d = use_density),
                            .f = bin_labeller, # bin_labeller is one of my helper functions
                            bin_data = hamby224_bin_boundaries,
                            case_info = hamby224_lime_cases)) %>%
    mutate(feature = factor(feature),
           nbins = factor(nbins),
           feature_number = readr::parse_number(feature_desc),
           strictly_less = FALSE) %>%
    arrange(nbins)

@@

  # Reorder the variables of feature_desc and feature_bin for plotting purposes
  hamby224_test_explain <- hamby224_test_explain %>%
    mutate(feature_desc = reorder(feature_desc, strictly_less),
           feature_desc = reorder(feature_desc, feature_number),
           feature_desc = reorder(feature_desc, as.numeric(feature))) %>%
    select(bin_continuous:use_density, case:feature_desc, feature_bin:strictly_less, 
           data, prediction)

@@

```{r}
hamby224_lime_comparisons <- hamby224_test_explain %>%
  select(-data, -prediction) %>%
  group_by(case, bin_continuous, quantile_bins, nbins, use_density) %>%
  slice(1) %>%
  ungroup() %>%
  select(bin_continuous, quantile_bins, nbins, use_density, case, set, rf_features, rfscore, model_prediction, model_r2) %>%
  mutate(diff = rfscore - model_prediction,
         mean = (rfscore + model_prediction) / 2)

hamby224_lime_comparisons %>%
  group_by(bin_continuous, quantile_bins, nbins, use_density, set) %>%
  summarise(mse = (sum(diff)^2) / length(diff)) %>%
  arrange(set)

hamby224_lime_comparisons %>%
  filter(nbins == "2", set == "Set 1") %>%
  select(rf_features, diff, -case) %>%
  gather(key = feature, value = feature_value, rf_features) %>%
  select(feature, feature_value, diff) %>%
  ggplot(aes(x = feature_value, y = diff)) + 
  geom_point() + 
  facet_wrap( ~ feature, scales = "free") + 
  geom_hline(yintercept = 0, color = "blue")
```

@@

```{r fig.width = 8, fig.height = 4, eval = FALSE}
ggplot(hamby224_lime_comparisons, aes(x = rfscore, y = model_prediction)) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid(nbins ~ set) + 
  theme_bw() + 
  stat_smooth(method = "lm")
```

@@

```{r fig.width = 8, fig.height = 4, eval = FALSE}
ggplot(hamby224_lime_comparisons, aes(x = rfscore, y = abs(diff))) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid(nbins ~ set) +
  theme_bw()
```

@@

```{r fig.width = 8, fig.height = 4, eval = FALSE}
ggplot(hamby224_lime_comparisons, aes(x = rfscore, y = diff)) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid(nbins ~ set) +
  geom_hline(yintercept = 0, color = "blue") +
  theme_bw()
```

@@

```{r, eval = FALSE}
hamby224_test_explain %>%
  ggplot(aes(x = feature_desc)) + 
  geom_bar() +
  coord_flip() + 
  facet_grid(set ~ nbins)
```

```{r, eval = FALSE}
hamby224_test_explain %>%
  filter(set == "Set 1", nbins == "3") %>%
  mutate(rfscore = round(rfscore, 3)) %>%
  select(case, bullet1, bullet2, land1, land2, rfscore) %>%
  distinct() %>%
  ggplot(aes(x = land1, y = land2, label = bullet1, label2 = bullet2,
             text = paste('Bullets Compared: ', bullet1, "-", land1, 
                          "vs", bullet2, "-", land2,
                          '\nRandom Forest Score: ', 
                          ifelse(is.na(rfscore), "Missing due to tank rash", rfscore)))) +
  geom_tile(aes(fill = rfscore)) +
  facet_grid(bullet2 ~ bullet1, scales = "free") +
  theme_minimal() +
  scale_fill_gradient2(low = "darkgrey", high = "darkorange", midpoint = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "RF Score")
```
